Applying Renormalization Group Theory to the Square Well Liquid

### 2014-05-27 papers/thesis-roth/project.tex

The \emph{configuration integral} (the potential energy component of the partition function) and its derivative with respect to $\lambda$ are
\begin{align}
  Z_N(\lambda) &= \int d\rr^N\, e^{-\beta V_N(\lambda)} \\
  \hookrightarrow \frac{\partial Z_N(\lambda)}{\partial\lambda} &=  \int d\rr^N\, \frac{\partial}{\partial\lambda}e^{-\beta V_N(\lambda)} \\
  &= \int d\rr^N\, -\beta\frac{\partial V_N(\lambda)}{\partial\lambda}e^{-\beta V_N(\lambda)} \\
  &= \int d\rr^N\, -\beta W_N e^{-\beta V_N(\lambda)}
\end{align}
And the non-ideal part of the free energy (i.e.~the excess free energy) and its derivative are
\begin{align}
  F &= -\kT\ln\left( \frac{Z_N}{V^N} \right) \\
  \hookrightarrow \frac{\partial F}{\partial\lambda} &= \frac{V^N}{Z_N}\frac{\partial}{\partial\lambda}\left( \frac{Z_N}{V^N} \right) \\
  &= -\kT\left( \frac{1}{Z_N}\frac{\partial Z_N}{\partial\lambda} \right) \\
  &= \frac{1}{Z_N} \int d\rr^N\, W_N e^{-\beta V_N(\lambda)}\\
  &= \left\langle W_N \right\rangle_\lambda
\end{align}
(where $-\kT$ and $-\beta$ have divided out to $1$). Here, $V^N$ is
the total volume taken to the $N^{th}$ power, and
$\left\langle W_N \right\rangle_\lambda$ is the ensemble
average of $W_N$, with the subscript indicating some given
value of $\lambda$. We can simplify this even further by noting
\begin{align}
  \left\langle W_N \right\rangle &= \left\langle \sum_i\sum_jw(\mathbf{r}_i,\mathbf{r}_j) \right\rangle \\
  &= \frac{1}{2}N^2\left\langle w(\1,\2) \right\rangle
  \intertext{which gives us}
  \frac{\partial F}{\partial\lambda} &= \frac{1}{2}N^2\left\langle w(\1,\2) \right\rangle_\lambda
\end{align}

At this point, note that we have an exact equation with no approximation. Now take a series expansion:
\begin{align}\begin{split} % too long; use split environment, and align to a space (&{})
  \left\langle W_N(\lambda) \right\rangle_\lambda = &{} \left\langle W_N(\lambda)\right\rangle_{\lambda = 0} + (\lambda)\frac{\partial}{\partial\lambda}\left\langle W_N(\lambda) \right\rangle_{\lambda}\bigg|_{\lambda = 0} \\ &{} + \mathcal{O}(\lambda^2)
\end{split}\end{align}
The derivative term is
\begin{align}
  \frac{\partial}{\partial\lambda}\left\langle W_N(\lambda) \right\rangle_{\lambda} &= \left\langle V_N^{''} \right\rangle_\lambda - \beta\sigma_\lambda^2
\end{align}
where
\begin{align}
  \sigma_\lambda^2 &\equiv \left\langle \left[ W_N(\lambda) \right]^2 \right\rangle_\lambda - \left\langle W_N(\lambda) \right\rangle^2_\lambda
\end{align}
Now note that $W_N$ has no dependence on $\lambda$, therefore $W_N' = 0$, and
\begin{align}
  \frac{\partial}{\partial\lambda}\left\langle W_N(\lambda) \right\rangle_{\lambda} &= - \beta\sigma_\lambda^2 \\
  \therefore \left\langle W_N(\lambda) \right\rangle_\lambda &= \left\langle W_N(\lambda)\right\rangle_{\lambda = 0} - \lambda\beta\sigma_{\lambda=0}^2 + \mathcal{O}(\lambda^2)
\end{align}

\davidsays{The derivative is (pretty) easy if you write your ensemble
  average as a multidimensional integral, as in
  Eq.~\ref{eqn:dFdlambda}.  Write down $dF/d\lambda$ in terms of $Z_N$
  and an integral of Boltzmann factors times $W_N$.  Then when you
  take a derivative, it hits top and bottom, but each is an integral
  of an exponential, so those are easy to differentiate.  A little
  math will give you this variance formula.}

\fixme{Below this, I have not updated}

Putting this together, we have
\begin{align}
  \begin{split}
    \beta F = &{} \beta F_0 + \beta(1 - 0)\left\langle W_N(\lambda) \right\rangle_{\lambda = 0} \\ &{} - \beta\frac{1}{2}(1 - 0)^2\left(\left\langle W_N'(\lambda = 0) \right\rangle_{\lambda = 0} - \beta^2\sigma^2_{\lambda = 0} \right)
  \end{split} \\
  \begin{split}
    =&{}  \beta F_0 + \beta\left\langle W_N(\lambda) \right\rangle_{\lambda = 0} \\ &{} - \beta\frac{1}{2}\left(\left\langle W_N'(\lambda = 0) \right\rangle_{\lambda = 0} - \beta^2\sigma^2_{\lambda = 0} \right) \label{eqn:complicated-HTE}
  \end{split}
\end{align}
Now, we can define the total perturbation energy as
\begin{align}
  W_N &= \sum_{i=1}^N\sum_{j>i}^N w_\lambda(\mathbf{r}_i,\mathbf{r}_j)
\end{align}
Then, following from equation~\ref{eqn:small-perturbation},
\begin{align}
  W_N &= \frac{\partial}{\partial\lambda}\left(\sum_{i=1}^N\sum_{j>i}^N \lambda w(\mathbf{r}_i,\mathbf{r}_j)\right) \nonumber \\
  &= W_N \\
  \therefore W_N' &= 0
\end{align}
\davidsays{Ah ha! Here you do what I mentioned earlier.  I'd tend to
  put this earlier, so you can throw away the $V_N''$ term sooner.
  And also talking about $W_N$ with this definition I would find nicer
  and clearer than all the $V_N'$.}  With this in mind, we can
simplify equation~\ref{eqn:complicated-HTE} to the
\emph{high-temperature expansion}, first derived by
Zwanzig:\cite{Zwanzig54}
\begin{align}
  \begin{split}
    \beta F = &{} \beta F_0 + \beta\left\langle W_N \right\rangle_0  \\ &{} - \frac{1}{2}\beta^2\left( \left\langle W_N^2 \right\rangle_0 - \left\langle W_N \right\rangle_0^2 \right) + \mathcal{O}(\beta^3)
  \end{split}
\end{align}
Zwanzig further showed that the $n^{th}$ term could be written in terms of the mean fluctuations $\left\langle \left[ W_N - \left\langle W_N\right\rangle_0 \right]^\nu \right\rangle_0$ with $\nu \leq n$.

>>>>

The \emph{configuration integral} (the potential energy component of the partition function) and its derivative with respect to $\lambda$ are
\begin{align}
  Z_N(\lambda) &= \int d\rr^N\, e^{-\beta V_N(\lambda)} \\
  \hookrightarrow \frac{\partial Z_N(\lambda)}{\partial\lambda} &=  \int d\rr^N\, \frac{\partial}{\partial\lambda}e^{-\beta V_N(\lambda)} \\
  &= \int d\rr^N\, -\beta\frac{\partial V_N(\lambda)}{\partial\lambda}e^{-\beta V_N(\lambda)} \\
  &= \int d\rr^N\, -\beta W_N e^{-\beta V_N(\lambda)}
\end{align}
And the non-ideal part of the free energy (i.e.~the excess free energy) and its derivative are
\begin{align}
  F &= -\kT\ln\left( \frac{Z_N}{V^N} \right) \\
  \hookrightarrow \frac{\partial F}{\partial\lambda} &= \frac{V^N}{Z_N}\frac{\partial}{\partial\lambda}\left( \frac{Z_N}{V^N} \right) \\
  &= -\kT\left( \frac{1}{Z_N}\frac{\partial Z_N}{\partial\lambda} \right) \\
  &= \frac{1}{Z_N} \int d\rr^N\, W_N e^{-\beta V_N(\lambda)}\\
  &= \left\langle W_N \right\rangle_\lambda
\end{align}
(where $-\kT$ and $-\beta$ have divided out to $1$). Here, $V^N$ is
the total volume taken to the $N^{th}$ power, and
$\left\langle W_N \right\rangle_\lambda$ is the ensemble
average of $W_N$, with the subscript indicating some given
value of $\lambda$.

\fixme{The following could probably go someplace else}
We can simplify this even further by noting
\begin{align}
  \left\langle W_N \right\rangle &= \left\langle \sum_i\sum_jw(\mathbf{r}_i,\mathbf{r}_j) \right\rangle \\
  &= \frac{1}{2}N^2\left\langle w(\1,\2) \right\rangle
  \intertext{which gives us}
  \frac{\partial F}{\partial\lambda} &= \frac{1}{2}N^2\left\langle w(\1,\2) \right\rangle_\lambda
\end{align}
\fixme{(end)}

We can integrate this expression over $F\in[F_0,F]$ and $\lambda\in[0,1]$ to obtain
\begin{align}
  F - F_0 &= (1-0)\left\langle W_N \right\rangle_\lambda \nonumber \\
  \hookrightarrow F &= F_0 + \left\langle W_N \right\rangle_\lambda \label{eqn:F-inTermsW}
\end{align}
At this point, note that we have an exact equation with no approximation. All we have done was take a derivative, simplify, and integrate. Now take a series expansion:
\begin{align}
  \left\langle W_N \right\rangle_\lambda = &{} \left\langle W_N\right\rangle_{\lambda = 0} + (\lambda)\frac{\partial}{\partial\lambda}\left\langle W_N \right\rangle_{\lambda}\bigg|_{\lambda = 0} \nonumber\\
  &{} + \mathcal{O}(\lambda^2) \label{eqn:Wn-expansion}
\end{align}
The derivative term is
\begin{align}
  \frac{\partial}{\partial\lambda}\left\langle W_N \right\rangle_{\lambda} =&{} \frac{\partial}{\partial\lambda}\left( \frac{1}{Z_N}\int d\rr^N\, W_N e^{-\beta V_N}\right) \\
  =&{} \frac{1}{Z_N}\frac{\partial}{\partial\lambda}\left( \int d\rr^N\, W_N e^{-\beta V_N} \right) \nonumber\\
  &{} - \left( \int d\rr^N\, W_N e^{-\beta V_N} \right)\frac{1}{Z_N^2}\frac{\partial}{\partial\lambda}Z_N \\
  =&{} \frac{1}{Z_N}\int d\rr^N\, \left( e^{-\beta V_N}\frac{\partial}{\partial\lambda}W_N + W_N\frac{\partial}{\partial\lambda}e^{-\beta V_N} \right) \nonumber \\
  &{} - \frac{1}{Z_N^2}\int d\rr^N\, \frac{\partial}{\partial\lambda}e^{-\beta V_N} \\
\end{align}
Note that $\partial/\partial\lambda W_N = 0$, so we have
\begin{align}
   \frac{\partial}{\partial\lambda}\left\langle W_N \right\rangle_{\lambda} =&{} \frac{1}{Z_N}\int d\rr^N (-\beta) W_N^2 e^{-\beta V_N} \nonumber\\
   &{} - \frac{1}{Z_N^2}\Bigg(\int d\rr^N\, W_N e^{-\beta V_N} \nonumber\\
   &{} \times \int d\rr^N\, (-\beta)W_Ne^{-\beta V_N}\Bigg) \\
   =&{} -\beta\Bigg[ \frac{1}{Z_N}\int d\rr^N\, W_N^2e^{-\beta V_N}  \nonumber\\
   &{} - \left( \frac{1}{Z_N}\int d\rr^N\, W_N e^{-\beta V_N} \right)^2 \Bigg] \\
   =&{} -\beta\left[ \left\langle W_N^2 \right\rangle_\lambda - \left\langle W_N \right\rangle_\lambda^2 \right] \\
   \equiv &{} -\beta\sigma_\lambda^2
\end{align}
Combining this result with equation~\ref{eqn:Wn-expansion} gives us
\begin{align}
  \left\langle W_N \right\rangle_\lambda &= \left\langle W_N\right\rangle_{\lambda = 0} - \beta\lambda\sigma_{\lambda=0}^2 + \mathcal{O}(\lambda^2) \label{eqn:Wn-expansion-simplified}
\end{align}
Putting together equations~\ref{eqn:F-inTermsW} and~\ref{eqn:Wn-expansion-simplified}, we have
\begin{align}
  F &= F_0 + \left\langle W_N \right\rangle_{\lambda=0} - \beta\lambda\sigma_{\lambda=0}^2 + \mathcal{O}(\lambda^2)
\end{align}
%% \begin{align}
%%   \begin{split}
%%     \beta F = &{} \beta F_0 + \beta(1 - 0)\left\langle W_N \right\rangle_{\lambda = 0} \\ &{} - \beta\frac{1}{2}(1 - 0)^2\left(\left\langle W_N'(\lambda = 0) \right\rangle_{\lambda = 0} - \beta^2\sigma^2_{\lambda = 0} \right)
%%   \end{split} \\
%%   \begin{split}
%%     =&{}  \beta F_0 + \beta\left\langle W_N \right\rangle_{\lambda = 0} \\ &{} - \beta\frac{1}{2}\left(\left\langle W_N'(\lambda = 0) \right\rangle_{\lambda = 0} - \beta^2\sigma^2_{\lambda = 0} \right) \label{eqn:complicated-HTE}
%%   \end{split}
%% \end{align}
This is the \emph{high-temperature expansion}, first derived by
Zwanzig:\cite{Zwanzig54}
\fixme{Zwanzig has a factor of $1/2$ that I'm missing}
\fixme{Below this point has not been updated}
\begin{align}
  \begin{split}
    \beta F = &{} \beta F_0 + \beta\left\langle W_N \right\rangle_0  \\ &{} - \frac{1}{2}\beta^2\left( \left\langle W_N^2 \right\rangle_0 - \left\langle W_N \right\rangle_0^2 \right) + \mathcal{O}(\beta^3)
  \end{split}
\end{align}
Zwanzig further showed that the $n^{th}$ term could be written in terms of the mean fluctuations $\left\langle \left[ W_N - \left\langle W_N\right\rangle_0 \right]^\nu \right\rangle_0$ with $\nu \leq n$.
